---
title: "Module5"
output: html_document
date: "2025-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Sources
Content for this Module is derived from
* https://rebeccabarter.com/blog/2020-03-25_machine_learning
* https://education.rstudio.com/blog/2020/02/conf20-intro-ml/
* https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/#google_vignette

ChatGPT was used in the creation of examples and explanations.

### Objectives 
By the end of the session, participants will be able to:
	* Understand key concepts in AI & machine learning.
	* Perform data preprocessing for ML tasks.
	* Train and evaluate basic models using caret, tidymodels, randomForest, and xgboost.
	* Interpret key model evaluation metrics.

### Introduction

AI versus Machine Learning

Machine learning is a subset of AI. "Artificial intelligence refers to the general ability of computers to emulate human thought and perform tasks in real-world environments, while machine learning refers to the technologies and algorithms that enable systems to identify patterns, make decisions, and improve themselves through experience and data." Deep learning is a type of machien learning using neural networks. You can read more:
* https://ai.engineering.columbia.edu/ai-vs-machine-learning/
* https://professionalprograms.mit.edu/blog/technology/machine-learning-vs-artificial-intelligence/
* https://www.ibm.com/think/topics/ai-vs-machine-learning-vs-deep-learning-vs-neural-networks

The original package for machine learning in R is `caret`: Caret stands for Classification And Regression Training. More recently, `tidymodels` uses the grammar of the tidyverse as a collection of modelling packages. `caret` is more widely used while `tidymodels` offers a modern, modular approach. 

```{r}

library(caret)
library(tidymodels)
library(tidyverse)
library(randomForest)
library(xgboost)
library(skimr)
library(janitor)

```

The first step is an exploratory data analysis—plots and summaries of the data to understand the variables and their distributions. However, we have done that in previous modules. 

### data split 
Training data is used for the model training and hyperparameter tuning. Once trained, the model can be evaluated against test data to assess accuracy.

```{r}
# split into training and testing datasets. Stratify by Sale price 
ames_split <- rsample::initial_split(
  ames_data, 
  prop = 0.2, 
  strata = sale_price
)

```

### Preprocessing

Common preprocessing tasks are to convert factors to numbers, handle missing data, and normalize/standardize where appropriate.

### Random Forest

### xgboost

XGBoost is robust against highly skewed and/or correlated data, so the amount of preprocessing required is minimal. 
