---
title: "Module3"
output: html_document
date: "2025-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Module 3: Data visualization with ggplot2

#### Sources
Content for this Module is derived from:
* https://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2.html

ChatGPT was used in the creation of examples and explanations.

### Objectives
By the end of this session, learners will be able to:
	*	Understand the grammar of graphics
	*	Create basic visualizations using ggplot2
	*	Customize aesthetics, labels, and themes
	*	Use facet_wrap() to explore subgroups
	
### Introduction

The "gg" in `ggplot2` stands for grammar of graphics. The plots are built in layers and you can additively string together functions with `+` to add layers and customize your plots.

```{r}

#read in the data we used in Module 2

username<-"YOUR USERNAME HERE"
path<-paste0("/mnt/work/workbench/",username,"/scratch/",username,"/data/")
path2<-paste0(path,"sample.csv")

dat<-read_csv(path2)

# Start with the dataset and aesthetic mapping
ggplot(data = dat, aes(x = Age, y = BloodPressure))

#Nothing appears — because no geometric layer is added yet!

# Add a geometric layer: scatter plot
ggplot(data = dat, aes(x = Age, y = BloodPressure)) +
  geom_point()

#the plot will show up on the bottom right of your RStudio IDE


```

# Common plots

Histograms visualize the distribution of a continuous variable by dividing its range into consecutive intervals called bins and then counting the number of observations that fall into each bin. By default, `geom_histogram()` has 30 bins. Binwidth controls the width of the bins (or bars) along the x-axis. The binwidth argument directly specifies the size of these intervals. For example, if your x-axis represents age in years and you set binwidth = 5, each bar in the histogram will represent a 5-year age range.

`ggplot` has many themes that allow for publication ready plots. You can tweak the display of existing themes or make your own.
Read more [here](https://ggplot2.tidyverse.org/reference/ggtheme.html). 

```{r}

# Let's make a histogram
ggplot(data = dat, aes(x = Age)) +
  geom_histogram()

# Let's customize the histogram with arguments for bin width and color
ggplot(data = dat, aes(x = Age)) +
  geom_histogram(binwidth=5,fill="blue") 


# Bar plot for categorical data; geom_bar uses counts by default; now we add a theme for a nicer presentation
ggplot(data=dat, aes(x = Group)) +
  geom_bar(fill = "darkorange") + theme_bw()

```


### Exporting the plot

The `ggsave()` function allows you to export a plot created with ggplot. You can specify the dimension and resolution of your plot by adjusting the appropriate arguments (width, height and dpi) to create high quality graphics for publication. 

I also like to use base-r functions `pdf()` and `png()`. These function serves as a graphics device driver, allowing you to direct plots generated in R to a PDF or png file. You use it in conjunction with dev.off() to open and then close the file for writing.

```{r}

#approach 1
my_plot<-
ggsave(filename = "test_plot.png", plot = my_plot, width = 12, height = 10, dpi = 300, units = "cm")

#approach 2
pdf("test_plot2.png",height=3,width=3,useDingbats=FALSE)
my_plot
dev.off()

?png


```

### Challenge

Task: Given a simple data frame of patients with ID, sex, age, and cholesterol values in two visits, write code to:
	1.	

```{r}



```


