---
title: "Module5"
output: html_document
date: "2025-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Sources
Content for this Module is derived from
* https://rebeccabarter.com/blog/2020-03-25_machine_learning
* https://education.rstudio.com/blog/2020/02/conf20-intro-ml/
* https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/#google_vignette

ChatGPT was used in the creation of examples and explanations.


### Introduction

AI versus Machine Learning
Machine learning is a subset of AI. "Artificial intelligence refers to the general ability of computers to emulate human thought and perform tasks in real-world environments, while machine learning refers to the technologies and algorithms that enable systems to identify patterns, make decisions, and improve themselves through experience and data." You can read more:
* https://ai.engineering.columbia.edu/ai-vs-machine-learning/
* https://professionalprograms.mit.edu/blog/technology/machine-learning-vs-artificial-intelligence/

The original package for machine learning in R is `caret`: Caret stands for Classification And Regression Training. More recently, `tidymodels` uses the grammar of the tidyverse as a collection of modelling packages.

```{r}

library(caret)
library(tidymodels)
```

The first step is an exploratory data analysis—plots and summaries of the data to understand the variables and their distributions. However, we have done that in previous modules. 

### data split 
Training data is used for the model training and hyperparameter tuning. Once trained, the model can be evaluated against test data to assess accuracy.

```{r}
# split into training and testing datasets. Stratify by Sale price 
ames_split <- rsample::initial_split(
  ames_data, 
  prop = 0.2, 
  strata = sale_price
)

```

### Preprocessing

### random forest

### xgboost

XGBoost is robust against highly skewed and/or correlated data, so the amount of preprocessing required is minimal. 
