---
title: "Module 2"
author: "Brooke Wolford"
date: "2025-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Module 2: Data Handling in R

This lesson is based on content from [Software Carpentry](https://swcarpentry.github.io/r-novice-inflammation/01-starting-with-data.html).

First, we need to load in the libraries we will use.

```{r}
#let's install a package
if(!requireNamespace(c("data.table","tidyverse"), quietly = TRUE)) {
  install.packages(c("data.table","tidyverse"))
}
#now we call the package
library(data.table)
library(tidyverse)
#The tidyverse package is an “umbrella-package” that installs several packages useful for data analysis which work together well such as tidyr, dplyr, ggplot2, tibble, etc.
```

Now, let's read in our data. Each row holds the observations for just one patient. Each column holds the inflammation measured in a day, so we have a set of values in successive days.

I like to use the data.table package which has a function `fread()`. However, you can also use base r function `read_csv()`. We will also inspect our data to get a sense of what it looks like.

```{r} 
#let's download data 
if(!file.exists("data/r-novice-inflammation-data.zip")){
  download.file("https://swcarpentry.github.io/r-novice-inflammation/data/r-novice-inflammation-data.zip", destfile = "data/r-novice-inflammation-data.zip", mode = "wb")
}
dat<-read.csv(unz("data/r-novice-inflammation-data.zip", "data/inflammation-01.csv"),header=FALSE)

#what do you think the header option does? how can we find out? what would happen if we did not use it?

head(dat)

dim(dat)

nrow(dat)

names(dat)

str(dat)

summary(dat)

```

Like many programming languages, we can use indices to specify a certain row and column. The first value in brackets is row, and the second is column. R starts indices at 1, but some other langauges use 0, so be mindful of this.

If we want to select more than one row or column, we can use the function c, which combines the values you give it into one vector or list.

If we want to select contiguous rows or columns, we can use the colon.

```{r}

#this will pull out the value in the 30th row and 20th column
dat[30,20]

dat[c(1, 3, 5), c(10, 20)]

#let's check what 1:5 does
1:5

#now let's use it in row and column space
dat[1:5,1:5]

```

If we leave the row or column space open, it means we wnt all of that row or column.

```{r}

#All columns from row 3
dat[3,]

# All rows from column 16-18
dat[, 16:18]

#what do you think happens if we leave both values empty
dat[,]

```

In tidyverse, we don't have to worry so much about indices. We can use column names

